Parsing Postfix log files

### 2009-03-30 doc/logparser-introduction.tex
Wrote introduction for implementation chapter.
Major work on the introduction chapter: the blurb at the start has been extended with content from the SGAI paper, and the layout section is more or less finished.

The architecture described herein was developed as part of a larger project
to improve anti-spam defences by analysing the performance of the set of
filters currently in use, optimising the order and membership of the set
based on that analysis, and developing supplemental filters where
deficiencies are identified.  Most anti-spam techniques are content-based
(e.g.~\cite{a-plan-for-spam, word-stemming, relaxed-online-svms}) and
require the mail to be accepted before determining if it is spam, but
rejecting mail during the delivery attempt is preferable: senders of
non-spam mail that is mistakenly rejected will receive an immediate
non-delivery notice; resource usage is reduced on the accepting mail server
(allowing more intensive content-based techniques to be used on the mail
that is accepted); users have less spam mail to wade through.  Improving
the performance of anti-spam techniques that are applied when mail is being
transferred via \acronym{SMTP} is the goal of this project, by providing a
platform for reasoning about anti-spam filters.  The approach chosen to
measure performance is to analyse the log files produced by the
\acronym{SMTP} server in use, Postfix, rather than modifying its source
code to generate statistics: this approach improves the chances of other
Postfix users testing and using the software developed for this project.
The need arose for a parser capable of dealing with the great number and
variety of log lines produced by Postfix: the parser must be designed so
that adding support for parsing new inputs is a simple task, because the
log lines to be parsed will change over time.

Most mail server administrators will have performed some basic processing
of the log files produced by their mail server at one time or another,
whether it was to debug a problem, explain to a user why their mail is
being rejected, or check if new anti-spam measures are working.  The more
adventurous will have generated statistics to show how successful each of
their anti-spam measures has been in the last week, and possibly even
generated some graphs to clearly illustrate these statistics to management
or users.\footnote{This was the first real foray the author, a Systems
Administrator for a network of over 2,000 computers and over 1,800 users,
took into processing Postfix log files.}  Very few will have performed
in-depth parsing and analysis of their log files, where the parsing must
correlate the log lines per-connection or per-queueid rather than
processing log lines independently.  One of the barriers to this kind of
processing is the unstructured nature of Postfix log files, where each log
line was added on an ad hoc basis as a requirement was discovered or new
functionality was added.\footnote{A history of all changes made to Postfix
is distributed with the source code, available from
\urlLastChecked{http://www.postfix.org/}{2009/02/23}.}  Further
complication arises because the set of rejection messages is not fixed: new
messages can be added by the administrator with custom checks; every
\acronym{DNSBL}\footnote{This thesis is supplied with a glossary
(\textsection\ref{Glossary}) and a list of acronyms
(\textsection\ref{Acronyms}).} returns a different explanatory message;
policy servers may log different messages depending on the characteristics
of the connection.  There are many ways in which the log lines may differ
between servers, even within the same organisation: servers may be
configured differently, or running different versions of Postfix.

This thesis documents the difficult process of parsing Postfix log files,
presenting an architecture designed to enable the users of a parser to
easily extend it to parse their particular inputs, without requiring a lot
of work or a high level of understanding of the parsing process and the
parser's internal workings.  This architecture is the basis of
\parsername{}, a program that parses Postfix log files and places the
resulting data into a database for later analysis.  The gathered data can
be used to optimise current anti-spam measures, provide a baseline to test
new anti-spam measures against, or to produce statistics showing how
effective those measures are.  Numerous other uses are possible for such
data: improving server performance by identifying troublesome destinations
and reconfiguring appropriately; identifying regular high volume uses
(e.g.\ customer newsletters) and restricting those uses to off-peak times
or providing a dedicated service for them; detecting virus outbreaks that
propagate via mail; as a base for billing customers on a shared server.
Preserving the raw data enables users to develop a multitude of uses far
beyond those conceived of by the author.

@@

Chapter~\ref{parser architecture} describes the parser architecture
developed for this project, beginning with an overview of the architecture,
then describing each of the components of the architecture in detail.

XXX THIS ALL NEEDS TO BE CHECKED AND UPDATED\@.

Chapter~\ref{Postfix Parser Implementation} discusses the parsing rules in
detail, explaining the purpose and usage of each field in a rule, referring
to an example rule and sample log line it successfully recognises.  The
pros and cons of overlapping rules are considered, including techniques for
detecting unintentional overlaps.  Rule efficiency concerns are discussed,
in particular the optimisations used by the algorithm.  The section
concludes with a description of using the tools provided with the parser to
generate new rules (specifically the regex in each rule) from unparsed log
lines.  It contains the core of the paper, describing a naive parsing
algorithm and the complications initially encountered that shaped the full
algorithm.  A flow chart and a discussion of the emergent behaviour
exhibited by the algorithm accompanies a comprehensive explanation of the
different stages of the initial algorithm.  The framework that actions and
rules fit into is documented, then the actions taken during execution of
the algorithm are described, followed by the process of adding a new
action.  The section concludes with an in-depth description of the further
complications discovered, and their solutions that complete the parser.
This algorithm requires a database for storing both the rules used when
parsing and the results gleaned from parsing.  The database schema used is
described in \sectionref{database}, explaining in detail the tables used
for storing the data gleaned from the log files and the table that stores
the rules.

Chapter~\ref{parsing coverage} analyses the coverage the parser achieves
over a set of \numberOFlogFILES{} log files taken from a mail server
handling mail for over 700 users, averaging 8500 mails per day
(\graphref{Mails received per day}).  Coverage is described both
in terms of the fraction of log lines parsed and the fraction of mails and
connections successfully reconstructed by the parsing algorithm; dealing
with false negatives and a discussion of the difficulties in identifying
false positives is also included.  As part of determining the coverage of
the parser a random sampling of log lines was parsed, and the correctness
of the results manually verified.

Chapter~\ref{conclusion} contains the conclusion of the thesis, describing
the results of the research, design, and implementation of the parser.

The bibliography (appendix~\ref{bibliography}) contains references to the
resources used in developing the algorithm, writing the program, and
preparing this thesis.

@@

Appendix~\ref{Acronyms} provides a list of acronyms used in the thesis.

Appendix~\ref{Postfix Daemons} gives a brief description of Postfix
daemons.

@@

Portions of \textsection\ref{introduction}, \textsection\ref{parser
architecture}, \textsection\ref{Postfix Parser Implementation}, and
\textsection\ref{Results} have previously been published at a
conference~\cite{sgai-2008}, and later reprinted in a
journal~\cite{elsevier-2009}.  Publication of the conference paper and
journal article was supported by Science Foundation Ireland
RFP~05/RF/CMS002.

>>>>

The architecture and implementation described in this thesis was developed
as the foundation of a larger project to improve anti-spam defences, by
analysing the performance of the set of filters currently in use,
optimising the order and membership of the set based on that analysis, and
developing supplemental filters where deficiencies are identified.  Most
anti-spam techniques are content-based (e.g.~\cite{a-plan-for-spam,
relaxed-online-svms, word-stemming}) and require the mail to be accepted
before determining if it is spam, but rejecting mail during the delivery
attempt is preferable: senders of non-spam mail that is mistakenly rejected
will receive an immediate non-delivery notice; resource usage is reduced on
the accepting mail server, allowing more intensive content-based techniques
to be used on the mail that is accepted; users have less spam mail to wade
through.  Improving the performance of anti-spam techniques that are
applied when mail is being transferred via \acronym{SMTP} is the goal of
this project, by providing a platform for reasoning about anti-spam
filters.

The approach chosen to measure performance is to analyse the log files
produced by the \acronym{SMTP} server in use, Postfix, rather than
modifying its source code to generate statistics: this approach improves
the chances of other Postfix users testing and using the software developed
for this project.  The need arose for a parser capable of dealing with the
great number and variety of log lines produced by Postfix: the parser must
be designed so that adding support for parsing new inputs is a simple task,
because the log lines to be parsed will change over time.  This variety in
log lines occurs for several reasons:

\begin{itemize}

    \squeezeitems{}

    \item Log lines differ amongst versions of Postfix.

    \item The mail administrator can define custom rejection messages.

    \item Policy servers~\cite{policy-servers} may log different messages
        depending on the characteristics of the connection.

    \item Every \acronym{DNSBL}\footnote{This thesis is supplied with a
        glossary (\textsection\ref{Glossary}) and a list of acronyms
        (\textsection\ref{Acronyms}).} returns a different explanatory
        message.

\end{itemize}

Most mail server administrators will have performed some basic processing
of the log files produced by their mail server at one time or another,
whether it was to debug a problem, explain to a user why their mail is
being rejected, or check if new anti-spam measures are working.  The more
adventurous will have generated statistics to show how successful each of
their anti-spam measures has been in the last week, and possibly even
generated some graphs to clearly illustrate these statistics to management
or users.\footnote{This was the first real foray the author, a Systems
Administrator for a network of over 2,000 computers and over 1,800 users,
took into processing Postfix log files.}  Very few will have performed
in-depth parsing and analysis of their log files, where the parsing must
correlate the log lines per-connection or per-queueid rather than
processing log lines independently.  One of the barriers to this kind of
processing is the unstructured nature of Postfix log files, where each log
line was added on an ad hoc basis as a requirement was discovered or new
functionality was added.\footnote{A history of all changes made to Postfix
is distributed with the source code, available from
\urlLastChecked{http://www.postfix.org/}{2009/02/23}.}  Further
complication arises because the set of rejection messages is not fixed, and
there are many ways in which the log lines may differ between servers, even
within the same organisation, where servers may be configured differently,
or even running different versions of Postfix.

It was hoped to reuse an existing parser rather than writing one from
scratch, but the effort required to adapt and improve an existing parser
was judged to be greater than the effort to write a new one, because the
techniques used by the existing parsers severely limited their potential:

\begin{itemize}

    \item Some ignored the majority of log lines, parsing specific log
        lines accurately, but without any provision for parsing new or
        similar log lines.

    \item Others sloppily parsed the majority of log lines, but were
        incapable of distinguishing between log lines of the same category,
        e.g.\ rejecting a mail delivery attempt.

    \item None of the reviewed parsers extracted enough data for the
        purposes of this project.

\end{itemize}

The only prior published work on the subject of parsing Postfix log files
that the author is aware of is \textit{Log Mail Analyser: Architecture and
Practical Utilizations\/}~\cite{log-mail-analyser}, which aims to extract
data from log files, correlate it, and present it in a form suitable for a
systems administrator to search using the myriad of standard Unix text
processing utilities already available.  A full state of the art review is
provided in \sectionref{state of the art review}, beginning with a review
of \textit{Log Mail Analyser}.

The architecture developed is conceptually simple: provide a few generic
functions (\textit{actions\/}), each capable of dealing with an entire
category of inputs (e.g.\ rejecting a mail delivery attempt), accompanied
by a multitude of precise patterns (\textit{rules\/}), each of which
recognises one input variant and only that variant (e.g.\ rejection
by a specific \acronym{DNSBL}).  This architecture is ideally suited to
parsing inputs where the input is not fully understood or does not conform
to a fixed grammar: the architecture warns about unparsed inputs and other
errors, but continues parsing as best it can, allowing the developer of a
new parser to decide which deficiencies are most important and require
immediate attention, rather than being forced to fix the first error that
arises.

This thesis documents the difficult process of parsing Postfix log files,
presenting an architecture designed to enable the users of a parser to
easily extend it to parse their particular inputs, without requiring a lot
of work or a high level of understanding of the parsing process and the
parser's internal workings.  This architecture is the basis of
\parsername{}, a program that parses Postfix log files and places the
resulting data into a database for later analysis.  The gathered data can
be used to optimise current anti-spam measures, provide a baseline to test
new anti-spam measures against, or to produce statistics showing how
effective those measures are.  Numerous other uses are possible for such
data: improving server performance by identifying troublesome destinations
and reconfiguring appropriately; identifying regular high volume uses
(e.g.\ customer newsletters) and restricting those uses to off-peak times,
or providing a dedicated service for them; detecting virus outbreaks that
propagate via mail; as a base for billing customers on a shared server.
Preserving the raw data enables users to develop a multitude of uses far
beyond those conceived of by the author.

@@

Chapter~\ref{parser architecture} describes the parser architecture
developed for this project, beginning with an overview, then describing
each of the components of the architecture in detail.  This chapter
concentrates on the abstract, theoretical, implementation-independent
aspects of the architecture; the practical aspects are deferred until
chapter~\ref{Postfix Parser Implementation}.

Chapter~\ref{Postfix Parser Implementation} documents \parsername{}, the
parser based on the architecture described in chapter~\ref{parser
architecture}.  The practical difficulties of implementing each of the
components of the architecture are described, accompanied by the many
complications encountered when parsing Postfix log files, and other
important details of the implementation.

Chapter~\ref{Results} evaluates \parsernames{} efficiency, discussing the
various optimisations implemented in the framework, and the effect they
have.  It also discusses the coverage achieved by \parsername{} over
\numberOFlogFILES{} log files, with separate discussions of the number of
log lines correctly recognised, and the number of connections and mails
whose journey through Postfix was correctly reconstructed.

Chapter~\ref{conclusion} contains the conclusion of the thesis.

The bibliography (appendix~\ref{bibliography}) contains references to the
resources used in designing the architecture and writing \parsername{}.

@@

Appendix~\ref{Acronyms} contains a list of acronyms used in the thesis;
uncommon acronyms will have an entry in the glossary too.

Appendix~\ref{Postfix Daemons} concludes this thesis with a brief
description of Postfix daemons.

@@

Portions of chapters~\ref{introduction} and~\ref{parser
architecture}--\ref{Results} have previously been published at a
conference~\cite{sgai-2008}, and later reprinted in a
journal~\cite{elsevier-2009}.  Publication of the conference paper was
supported by Science Foundation Ireland RFP~05/RF/CMS002.

### 2009-03-30 doc/logparser.tex
Changes from proofreading, and a couple of other checks.
Added a \action{} command to have consistent formatting of action names. Changes \parsernames{} to have an ' before the s.
Dealt with some XXX items.
Changes made after reading my paper.
Corrections after proofreading.
Major work on the introduction chapter: the blurb at the start has been extended with content from the SGAI paper, and the layout section is more or less finished.

% Sort numbers where there are multiple citations.  Does not appear to have
% the expected effect (probably clashes with varioref or hyperref), though
% it does reduce the space between numbers.
\usepackage{cite}
% Check for unused references, and how the citation key (e.g.\ slct-paper)
% in the margin beside the reference.
%\usepackage[nomsgs]{refcheck}
% Show where references are used; neither work.
%\usepackage{citeref}
%\usepackage{backref}
%\renewcommand{\refname}{Bibliography}

@@

% \showgraph{filename}{caption}{label}
\newcommand{\showgraph}[3]{%
    % XXX REVISIT THE PLACEMENT OF FIGURES, THOUGH I THINK LATEX'S DEFAULTS
    % AR EPROBABLY PRETTY GOOD.
    \begin{figure}%[hbt!]
        \caption{#2}\label{#3}
        \includegraphics{#1}
    \end{figure}
}

%\showtable{filename}{caption}{label}
\newcommand{\showtable}[3]{
    \begin{table}[ht]
        \caption{#2}\label{#3}
        \input{#1}
    \end{table}
}

@@

\newcommand{\flowchart}[2]{%
    (action: #1\@; flowchart:~#2)%
}

@@

\newcommand{\numberOFlogFILESall}[0]{%
    522%
}

@@

% The name of the program, so I only have to change it in one place.
\newcommand{\parsername}{\acronym{PLP}}
\newcommand{\parsernames}{\acronyms{PLP}}
\newcommand{\parsernamelong}{Postfix Log Parser}

@@

@@

@@

    XXX ADD A $\backslash{}$action\{\} COMMAND TO FORMAT ACTION NAMES
    CONSISTENTLY\@.

@@

    XXX PLPs SHOULD BE PLP's IN A LOT OF PLACES\@; ADD A NEW COMMAND AND
    CHECK THE CURRENT USAGE\@.

@@

>>>>

% Sort numbers where there are multiple citations.  Does not appear to have
% the expected effect (probably clashes with hyperref), though it does
% reduce the space between numbers.
\usepackage{cite}
% Check for unused references, and how the citation key (e.g.\ slct-paper)
% in the margin beside the reference.
%\usepackage[nomsgs]{refcheck}
% Show where references are used; neither work.
%\usepackage{citeref}
%\usepackage{backref}
%\renewcommand{\refname}{Bibliography}

@@

% \showgraph{filename}{caption}{label}
\newcommand{\showgraph}[3]{%
    % XXX REVISIT THE PLACEMENT OF FIGURES AND TABLES, THOUGH I THINK
    % LATEX'S DEFAULTS ARE PROBABLY PRETTY GOOD.
    \begin{figure}%[hbt!]
        \caption{#2}\label{#3}
        \includegraphics{#1}
    \end{figure}
}

%\showtable{filename}{caption}{label}
\newcommand{\showtable}[3]{
    \begin{table}%[ht]
        \caption{#2}\label{#3}
        \input{#1}
    \end{table}
}

@@

\newcommand{\flowchart}[2]{%
    (action: \action{#1}\@; flowchart:~#2)%
}

\newcommand{\action}[1]{%
    \texttt{#1}%
}

@@

% XXX UPDATE THIS WHEN all-logs PASSES
\newcommand{\numberOFlogFILESall}[0]{%
    522%
}

@@

% The name of the program, so I only have to change it in one place.
\newcommand{\parsername}{\acronym{PLP}}
\newcommand{\parsernames}{\acronym['s]{PLP}}
\newcommand{\parsernamelong}{Postfix Log Parser}

@@

    XXX REWRITE WHEN THESIS IS FINISHED\@.

@@

    XXX ABSTRACT FROM SGAI PAPER\@.

    Some parsers need to be very precise and strict when parsing, yet must
    allow users to easily adapt or extend the parser to parse new inputs,
    without requiring that the user have an in-depth knowledge and
    understanding of the parser's internal workings.  This paper presents a
    novel parsing architecture, designed for parsing Postfix log files,
    that aims to make the process of parsing new inputs as simple as
    possible, enabling users to trivially add new rules (to parse variants
    of existing inputs) and relatively easily add new actions (to process a
    previously unknown category of input).  The architecture scales
    linearly or better as the number of rules and size of input increases,
    making it suitable for parsing large corpora or months of accumulated
    data.

@@

@@

@@

    XXX MAIL MADS ABOUT EXAMINERS\@.
